Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2022-11-08 paper/ch_potentials.tex
prep for final submission

% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\documentclass[manuscript,review]{acmart}

@@

\usepackage{todonotes}
% \usepackage{booktabs}
\usepackage{multirow}
\usepackage[algo2e,vlined]{algorithm2e}
\usepackage{todonotes}

@@

\begin{figure}
\centering
\includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
\caption{Vertices explored by A*. The color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
\label{img:search-space}
\Description[A map where vertices visited by A* are marked.]{
A map of Karlsruhe where vertices visited by A* are marked.
The shape of the area with the visited vertices is similar to a drop.
The source vertex roughly lies in the center of the drop, the target vertex at the tip.
}
\end{figure}

@@

\draw[dashed] (4) -- (6);
\draw[dashed] (6) -- (8);
\end{tikzpicture}
\caption{
Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$.
}
\label{fig:ch}
\Description[Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
The y-position on the figure indicates the rank of each vertex.
Where the original shortest path down and then up, a shortcut edge connects the two higher vertices.
The path in the augmented graph goes only up and then down.
}
\end{figure}

@@

\emph{Contraction Hierarchies} (CH) is a two-phase speedup technique to accelerate shortest path computations on road networks through precomputation.
For a detailed discussion we refer to~\cite{gssv-erlrn-12}.
Here, we only briefly introduce the necessary notation and algorithms used in this paper.
In a preprocessing phase, vertices are ordered totally by ``importance'' where more important vertices should lie on more shortest paths.
Intuitively, vertices on highways are more important than vertices on rural streets.
For CH, such an ordering is obtained heuristically.
The position of a vertex in the order is also denoted as its \emph{rank}.
Vertices of higher rank are informally referred to and visualized as ``higher up'' in the hierarchy.
Therefore, an edge where the tail has a lower rank than the head is an \emph{upward} edge.
Analogously, when the head vertex has the lower rank, the edge is said to go \emph{downward}.
Once such an importance order was obtained, all vertices are contracted successively by ascending importance.
To \emph{contract} a vertex means temporarily removing it from the graph while inserting \emph{shortcut edges} between more important neighbours to preserve shortest distances among them.
The result is an \emph{augmented graph} $G^+$ of original edges and shortcuts with weights $w^+$.
We often refer to the augmented graph split into an upward graph $\gchu = (V, \echu)$ containing only upward edges and a downward graph $\gchd = (V, \echd)$ containing only downward edges.

@@

Unless stated differently, for all bidirectional A* variants, we always alternate between removing a vertex from the forward and the backward queues.
We also evaluate expanding the search with the smaller minimum queue key in our experiments.
While this may sound sensible, our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.

@@

We consider a variety of different applications with slightly different problem models.
The goal is always to answer many shortest path queries quickly.
To describe our framework, we establish a shared notation:
Input to each query are vertices $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
In the simplest case, $w_q$ will be scalar edge weights.
Live traffic is an example of this.
The challenge in this scenario is that values of $w_q$ might change between queries.
However, $w_q$ can also represent something more complex than scalar numbers.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.
In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time, and the query takes an additional departure time parameter.

@@

To support only live traffic, we set $w_\ell$ to the free-flow travel time.
$w_q$ is set to the travel time accounting for current traffic.
As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
In a real-world application, values from $w_q$ could be updated between queries.
This is all that is necessary to apply CH-Potentials in a live traffic scenario.

@@

Alternatively, live traffic can be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
In this case, we use the CCH instead of the CH preprocessing to build the augmented graph.
This allows us to introduce updates to $w_\ell$ and the augmented graph within a few seconds using the CCH customization phase.
This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic} where live traffic was represented only in $w_q$.
In contrast, with CCH-Potentials, changes to the live traffic situation are handled by changing $w_\ell$ via customization.
Additional features, such as turn restrictions, can be implemented by adjusting $w_q$.
For severe traffic events, the three-phase approach will have significantly faster query running times than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it comes at the expense of a more complex setup and updating $w_\ell$ takes a few seconds.
% Also, the customization is not fast enough to rerun it for every query.
% Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.

>>>>

% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\documentclass[manuscript]{acmart}

@@

% \usepackage{booktabs}
\usepackage{multirow}
\usepackage[algo2e,vlined]{algorithm2e}

@@

\begin{figure}
\centering
\includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
\caption{Vertices explored by A*. The color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last. Background by \href{https://www.openstreetmap.org/copyright}{OpenStreeMap}.}
\label{img:search-space}
\Description[A map where vertices visited by A* are marked.]{
A map of Karlsruhe where vertices visited by A* are marked.
The shape of the area with the visited vertices is similar to a drop.
The source vertex roughly lies in the center of the drop, the target vertex at the tip.
}
\end{figure}

@@

\draw[dashed] (4) -- (6);
\draw[dashed] (6) -- (8);
\end{tikzpicture}
\caption{
Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down path in $G^+$.
}
\label{fig:ch}
\Description[Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
The y-position on the figure indicates the rank of each vertex.
Where the original shortest path down and then up, a shortcut edge connects the two higher vertices.
The path in the augmented graph goes only up and then down.
}
\end{figure}

@@

\emph{Contraction Hierarchies} (CH) is a two-phase speedup technique to accelerate shortest path computations on road networks through precomputation.
For a detailed discussion we refer to~\cite{gssv-erlrn-12}.
% Here, we only briefly introduce the necessary notation and algorithms used in this paper.
In a preprocessing phase, vertices are ordered totally by ``importance'' where more important vertices should lie on more shortest paths.
Intuitively, vertices on highways are more important than vertices on rural streets.
For CH, such an ordering is obtained heuristically.
The position of a vertex in the order is also denoted as its \emph{rank}.
Vertices of higher rank are informally referred to and visualized as ``higher up'' in the hierarchy.
Therefore, an edge where the tail has a lower rank than the head is an \emph{upward} edge.
Analogously, when the head vertex has the lower rank, the edge is said to go \emph{downward}.
Once such an importance order was obtained, all vertices are contracted successively by ascending importance.
To \emph{contract} a vertex means temporarily removing it from the graph while inserting \emph{shortcut edges} between more important neighbours to preserve shortest distances among them.
The result is an \emph{augmented graph} $G^+$ of original edges and shortcuts with weights $w^+$.
We often refer to the augmented graph split into an upward graph $\gchu = (V, \echu)$ containing only upward edges and a downward graph $\gchd = (V, \echd)$ containing only downward edges.

@@

Unless stated differently, for bidirectional A*, we always alternate between removing a vertex from the forward and the backward queues.
We also evaluate expanding the search with the smaller minimum queue key in our experiments.
While this may sound sensible, our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.

@@

We consider a variety of different applications with slightly different problem models.
The goal is always to answer many shortest path queries quickly.
To describe our framework, we establish a shared notation:
Input to each query are vertices $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
In the simplest case, $w_q$ will be scalar edge weights.
Live traffic is an example of this.
The challenge in this scenario is that values of $w_q$ might change between queries.
However, $w_q$ can also represent something more complex than scalar numbers.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.
In the case of traffic predictions, $w_q$ is a function mapping the edge entry time to the traversal time, and the query takes an additional departure time parameter.

@@

To support only live traffic, we set $w_\ell$ to the free-flow travel time.
$w_q$ is set to the travel time accounting for current traffic.
As traffic only increases travel times, $w_\ell$ is a valid lower bound for $w_q$.
In a real-world application, values from $w_q$ could be updated between queries.
This is all that is necessary to apply CH-Potentials in a live traffic scenario.

@@

Alternatively, live traffic can be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
In this case, we use the CCH instead of the CH preprocessing to build the augmented graph.
This allows us to introduce updates to $w_\ell$ and the augmented graph within a few seconds using the CCH customization phase.
This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic} where live traffic was represented only in $w_q$.
In contrast, with CCH-Potentials, changes to the live traffic situation are handled by changing $w_\ell$ via customization.
Additional features, such as turn restrictions, can be implemented by adjusting $w_q$.
For severe traffic events, the three-phase approach will have significantly faster query running times than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it leads to a more complex setup and updating $w_\ell$ takes a few seconds.
% Also, the customization is not fast enough to rerun it for every query.
% Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.
